Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-15 paper/sheaf_lics.tex
more on modalities

\newcommand{\modal}{\ensuremath{\ocircle}}
\newcommand \True {\mathrm{true}}
\newcommand \False {\mathrm{false}}
\newcommand \closure[1] {\overline{#1}}
\newcommand \Char[1] {\chi_{#1}}%{\mathrm{char}(#1)}
\newcommand \E {\mathcal{E}}
\newcommand \Hom[1] {\mathrm{Hom}_{#1}}
\newcommand \Sh[1] {\mathrm{Sh}_{#1}}
\newcommand \squash[1] {\| #1 \| }
\newcommand \fib[2] {\mathrm{fib}_{#1}(#2)}
\newcommand \colim[1] {\mathrm{colim}(Q)}

@@

Finally, one of the main difficulty in our formalization is that many
equalities which are trival (by proof irrelevance) in the topos theoretic
setting becomes relevant and requires a carreful use of univalence
with appropriate isomorphisms.

@@

\cite{shulman-higher-modalities}

\subsection{Definition}

We use the following definition of truncated left exact modalities~:
\begin{defi}
  Let $p$ be a truncation index. Then a left exact modality is the
  data of
  \begin{enumerate}[(i)]
  \item A predicate $P:\Type_p \to \HProp$
  \item For every $p$-truncated type $A$, a $p$-truncated type
    $\modal A$ such that $P(\modal A)$
  \item For every $p$-truncated type $A$, a map $\eta_A:A \to
    \modal A$
  \end{enumerate}
  such that
  \begin{enumerate}[(i)]
    \setcounter{enumi}{3}
  \item For every $p$-truncated types $A$ and $B$, if $P(B)$ then
    $$\left\{
      \begin{array}{rcl}
        (\modal A \to B) & \to & (A \to B) \\
        f & \mapsto & f \circ \eta_A
      \end{array} \right.$$
    is an equivalence.
  \item for any $A:\Type_p$ and $B:A \to \Type_p$ such that $P(A)$
    and $\prod_{x:A} P(B x)$, then $P\left( \sum_{x:A} B(x)\right)$
  \item for any $A:\Type_p$ and $x,y:A$, if $\modal A$ is
    contractible, then $\modal (x=y)$ is contractible.
  \end{enumerate}
  Conditions (i) to (iv) define a {\em reflective subuniverse}, (i) to
  (v) a {\em modality}.
\end{defi}

The type of all $p$-types such that $P$ will be noted $\Type^\modal_p$

Since basic operations (dependent products, products, sigma types) are
stable for truncation levels, all theorem in~\cite{hottbook}, chapter
7.7 remains true.
In particular, if $X:\Type_p$ verifies $P(X)$, then for any $x,y:X$, $P(x=y)$.

Another useful property is that we can extend property {\it (iii)} to
dependent product.

Moreover, left exactness implies in particular homotopy fibers preservations~:
\begin{prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$~:
$$\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X} (\modal f x = \eta_Y y).$$

Moreover the following diagram commute
$$\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) &
}$$
where $\pi_1 \circ \gamma = \eta_X$, and $\pi_2 \circ \gamma$ is the
usual modalisation of paths.
\end{prop}

As we want, a left exact modality preserves $\HProp$~:
\begin{prop}
  If $P:\HProp$, then $\modal \widehat P : \HProp$, where $\widehat P$
  is $P$ seen as a $p$-type.
\end{prop}

\subsection{Examples of Left Exact Modalities}

\subsection{The new type theory arising from a left exact modality}

@@

Consistency issues.

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification/}.

>>>>

\newcommand{\modal}{\ensuremath{\ocircle}}
\newcommand \True {\mathrm{true}}
\newcommand \False {\mathrm{false}}
\newcommand \closure[1] {\overline{#1}}
\newcommand \Char[1] {\chi_{#1}}%{\mathrm{char}(#1)}
\newcommand \E {\mathcal{E}}
\newcommand \Hom[1] {\mathrm{Hom}_{#1}}
\newcommand \Sh[1] {\mathrm{Sh}_{#1}}
\newcommand \squash[1] {\| #1 \| }
\newcommand \fib[2] {\mathrm{fib}_{#1}(#2)}
\newcommand \colim[1] {\mathrm{colim}(Q)}
\newcommand \zero {\mathbf{0}}
\newcommand \one {\mathbf{1}}

@@

Finally, one of the main difficulty in our formalization is that many
equalities which are trival (by proof irrelevance) in the topos theoretic
setting becomes relevant and requires a carreful use of univalence
with appropriate isomorphisms and a lot of reasoning on equalities
between paths.

@@

This section is devoted to the definition and analyze of left exact
modalities as they have been introduced in \cite{hottbook} and
\cite{shulman-higher-modalities}. Our definition differs slightly from
those work in that we restrict the definition of the modality to be
$n$-truncated types for a given truncation index $n$, instead of on
all types.
%
This implies that we cannot reuse directly all results developed
in~\cite{hottbook} as some care must be taken to check that truncation
levels are preserved.

Note that similar notion of modal operator have been already studied
extensively in non-dependent type
theory~\cite{benton1998computational}. In those works, they connect
modalities to computational monad in programming
languages~\cite{moggi-monad}. However, in our setting, modalities
correspond to idempotent monads, a property is rarely true 
in functional programming, except for the identity monad.

The notion of left exact modalities plays a central role in this paper
because Lawvere-Tierney topologies corresponds exactly to left exact
modalities at $-1$ and sheafification is described as an inductive
process that produces a left exact modality at level $n+1$ from a
level exact modality at level $n$.

Section~\ref{sec:definition} presents the definition and basic
properties of left exact modalities. 
%
Section~\ref{sec:new-type-theory} discusses how a left exact modality
induces a ne homotopy theory, and in which case this theory is
consistent.
%
Finally, Section~\ref{sec:examples-left-exact} introduces some
example, and in particular the modality for the dense topology.

\subsection{Definition and Basic Properties}
\label{sec:definition}

% We use the following definition of truncated left exact modalities:
\begin{defi}
  \label{sec:defin-basic-prop-1}
  Let $n\geq -1$ be a truncation index. A left exact modality at level
  $n$ is the data of
  \begin{enumerate}[(i)]
  \item A predicate $P:\Type_p \to \HProp$
  \item For every $n$-truncated type $A$, a $n$-truncated type
    $\modal A$ such that $P(\modal A)$
  \item For every $n$-truncated type $A$, a map $\eta_A:A \to
    \modal A$
  \end{enumerate}
  such that
  \begin{enumerate}[(i)]
    \setcounter{enumi}{3}
  \item For every $n$-truncated types $A$ and $B$, if $P(B)$ then
    $$\left\{
      \begin{array}{rcl}
        (\modal A \to B) & \to & (A \to B) \\
        f & \mapsto & f \circ \eta_A
      \end{array} \right.$$
    is an equivalence.
  \item for any $A:\Type_p$ and $B:A \to \Type_p$ such that $P(A)$
    and $\prod_{x:A} P(B x)$, then $P\left( \sum_{x:A} B(x)\right)$
  \item for any $A:\Type_p$ and $x,y:A$, if $\modal A$ is
    contractible, then $\modal (x=y)$ is contractible.
  \end{enumerate}
  Conditions (i) to (iv) define a {\em reflective subuniverse}, (i) to
  (v) a {\em modality}.
\end{defi}

The type of all modal $n$-types, \ie types for which $P$ holds, is
defined as %$\Type^\modal_p$, so 
$$
\Type^\modal_p \ \defeq \sum_{T : \Type_n} P (T).
$$
%
Since basic type formers (dependent products, products, sigma types)
preserves truncation levels, we can leverage all theorems on modalites
presented in~\cite[Chapter 7.7]{hottbook}.

We now summarize the main properties that are used in our
formalization.

\begin{prop}

stability properties
\vspace{1em}

\nt{state precisely some properties here, at least the 4 below}

\vspace{1em}

\begin{itemize}

\item
$\modal \one =\one $
\item
Stability under dependent product (7.7.2 of \cite{hottbook})
\item
 In particular, if $X:\Type_p$ verifies
$P(X)$, then for any $x,y:X$, $P(x=y)$.
\item
Another useful property is that we can extend property {\it (iii)} to
dependent product.
\end{itemize}
\end{prop}

As the definition of sheafification is based on subobject classifiers
and thus on homotopy fibers, we need to look at how left exact
modalities preserves homotopy fibers.
\begin{prop}
\label{sec:defin-basic-prop}
For any $n$-truncated types $X$ and $Y$,
and any map $f:X \to Y$, the modalisation of fiber of $f$ above any element $y:Y$
is the fiber of $\modal f$ above $\eta_Y y$~:
$$\modal \left(\sum_{x:X}  (f x = y)\right) = \sum_{x:\modal X} (\modal f x = \eta_Y y).$$

Moreover the following diagram commute
$$\xymatrix{
  \sum_{x:X} (f x = y) \ar[r]^\eta \ar[d]_\gamma & \modal \left(\sum_{x:X}  (f x = y)\right) \ar@{=}[dl]\\
  \sum_{x:\modal X} (\modal f x = \eta_Y y) &
}$$
where $\pi_1 \circ \gamma = \eta_X \circ \pi_1$, and $\pi_2 \circ
\gamma$ is given by the modalisation of paths.
\end{prop}
\begin{proof}
\nt{ça a du sens de mettre une preuve, non ? c'est non trivial et
  nouveau par rapport aux travaux sur les modalités.}
\end{proof}

In the same way, left exact modalities preserve homotopy types.
\begin{prop}
  Let $k \leq n$.
  If $P:\Type_k$, then $\modal \widehat P : \Type_k$, where $\widehat P$
  is $P$ seen as a $n$-type.
\end{prop}
\begin{proof}
  An $n$-truncated type $P$ can equivalently be described as a type for
  which the unique map to $\one$ is with $n$-truncated fibers. Thus, the
  property is a direct corollary of
  Proposition~\ref{sec:defin-basic-prop} and the fact that $\modal \one =
  \one$.
\end{proof}

\subsection{The new type theory arising from a left exact modality}
\label{sec:new-type-theory}

As noticed in \cite{hottbook}, left exact modalities corresponds 
to sub-$(\infty,1)$-topoi.

@@

\nt{state the property on IsEquiv to show preservation of univalence}

Every left exact modality induces a new type theory, but with more
properties, that can make it inconsistent. We now provide a simple
characterization of left exact modalities that give rise to a
consistent theory.

\begin{prop}
  A left exact modality $\modal$ induces a consistent type theory if
  and only if $\modal \zero$ can not be inhabited in the initial type
  theory. In that case, the modality is said to be consistent.
\end{prop}
\begin{proof}
  By condition (iv) of Definition~\ref{sec:defin-basic-prop-1},
  $\modal \zero$ is an initial object of $\Type^\modal_p$, and thus
  corresponds to false for modal mere proposition.
  % 
  As $\modal \one = \one$, $\Type^\modal_p$ is consistent when
  $\modal \zero \neq \one$, that is when there is no proof of
  $\modal \zero$.
\end{proof}

\subsection{Examples of Left Exact Modalities}
\label{sec:examples-left-exact}

@@

A Coq formalization is available at
\url{https://github.com/KevinQuirin/sheafification/}.
%
It is based on the homotopy type theory library available at
\url{https://github.com/HoTT/HoTT}
